## Overview

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible 
to collect a large amount of data about personal activity relatively 
inexpensively. These type of devices are part of the quantified self movement
- a group of enthusiasts who take measurements about themselves regularly to 
improve their health, to find patterns in their behavior, or because they are 
tech geeks. One thing that people regularly do is quantify how much of a 
particular activity they do, but they rarely quantify how well they do it. 
In this project, your goal will be to use data from accelerometers on the belt, 
forearm, arm, and dumbell of 6 participants. They were asked to perform 
barbell lifts correctly and incorrectly in 5 different ways.

The goal of this project is to build a machine learning algorithm to predict 
activity quality (classe) from activity monitors.


## Creating a prediction model

Loading data

```{r}
library(caret)

## Loading required package: lattice
## Loading required package: ggplot2

library(ggplot2)
library(randomForest)

## randomForest 4.6-12
## Type rfNews() to see new features/changes/bug fixes.
```

Now I will go ahead and download both the training and testing data files and take a look at the data provided to build our model. The goal of the model is to use any variables provided to predict the manner in which a person did the exercise (classe).

```{r}
#read in training and testing data

train <- read.csv("training.csv", na.strings=c("NA","#DIV/0!",""))
test <- read.csv("testing.csv", na.strings=c("NA","#DIV/0!",""))

names(train)
str(train)
summary(train)
summary(train$classe)#this is the outcome we want to predict
```

Split training/testing data

Before we do anything, we will set aside a subset of our training data for cross validation (40%).



```{r}
#we want to predict the 'classe' variable using any other variable to predict with


inTrain <- createDataPartition(y=train$classe, p=0.6, list=FALSE)
myTrain <- train[inTrain, ]
myTest <- train[-inTrain, ]
dim(myTrain)
dim(myTest)

#some exploratory plots
#featurePlot(x=train[, 150:159], y = train$classe, plot = 'pairs')

```


Feature selection

Now we can tranform the data to only include the variables we will need to build our model. We will remove variables with near zero variance, variables with mostly missing data, and variables that are obviously not useful as predictors.

```{r}
#first we will remove variables with mostly NAs (use threshold of >75%)
mytrain_SUB <- myTrain
for (i in 1:length(myTrain)) {
  if (sum(is.na(myTrain[ , i])) / nrow(myTrain) >= .75) {
    for (j in 1:length(mytrain_SUB)) {
      if (length(grep(names(myTrain[i]), names(mytrain_SUB)[j]))==1) {
        mytrain_SUB <- mytrain_SUB[ , -j]
      }
    }
  }
}

dim(mytrain_SUB)

#names(mytrain_SUB)

#remove columns that are obviously not predictors
mytrain_SUB2 <- mytrain_SUB[,8:length(mytrain_SUB)]

#remove variables with near zero variance
NZV <- nearZeroVar(mytrain_SUB2, saveMetrics = TRUE)
NZV #all false, none to remov

keep <- names(mytrain_SUB2)

```

Random Forest Model

I decided to use the random forest model to build my machine learning algorithm as it is appropriate for a classification problem as we have and based on information provided in class lectures this model tends to be more accurate than some other classification models.

Below I fit my model on my training data and then use my model to predict classe on my subset of data used for cross validation.

```{r}
#fit model - RANDOM FOREST
library(caret)
set.seed(223)

modFit <- randomForest(classe~., data = mytrain_SUB2)
print(modFit)

predict1 <- predict(modFit, myTest, type = "class")
confusionMatrix(myTest$classe, predict1)

predict_train <- predict(modFit, myTrain, type = "class")
confusionMatrix(myTrain$classe, predict_train)
```

Error

As we can see from the model summaries above, when we run the model on our test data for cross validation we get an accuracy of 99.4% that we can estimate to be our out of sample error. When the model is fitted to the training data used to build the model it shows 100% accuracy, which we can assume as our in sample error.

Apply to final test set

Finally, we apply our model to the final test data. Upon submission all predictions were correct!

```{r}
predict_FINAL <- predict(modFit, test, type = "class")
print(predict_FINAL)
##  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 
##  B  A  B  A  A  E  D  B  A  A  B  C  B  A  E  E  A  B  B  B 
## Levels: A B C D E
pml_write_files = function(x) {
  n = length(x)
  for (i in 1:n) {
    filename = paste0("problem_id_", i, ".txt")
    write.table(x[i], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)
  }
}

pml_write_files(predict_FINAL)
```